Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-05 paper/paper.qmd
Minor tweaks

There is not currently a consensus definition of CDI severity.
Some scoring systems leverage clinical data available during the course of CDI,
while others focus on adverse outcomes of CDI at 30 days after diagnosis
[@ressler_defining_2021].
We explored four different ways to define CDI cases as severe or not (@fig-flowchart).
The "IDSA" definition of severe CDI is based on laboratory values collected on
the day of diagnosis, with a case being severe if serum creatinine level is
greater than or equal to $1.5 mg/dL$ and the white blood cell count is greater
than or equal to $15 k/\mu L$ [@mcdonald_clinical_2018].
However, the IDSA score is known to be a poor predictor of adverse outcomes
[@stevens_validation_2020], although it is straightforward to collect.
The remaining definitions we employed focus on the occurrence of adverse
outcomes, which may be more clinically relevant.
The "attributable" severity definition is based on disease-related complications
defined by the CDC, where an adverse event of ICU admission, colectomy, or death
occurs within 30 days of CDI diagnosis, and the adverse event is determined to
be attributable to the CDI by physician chart review [@mcdonald_recommendations_2007].
However, physician chart review is time-consuming and has not been completed for
all cases, so we defined "all-cause" severity where a case is severe if an
adverse event occurs within 30 days of the diagnosis regardless of the cause of
the adverse event.
Finally, we defined a "pragmatic" severity definition that makes use of the
attributable definition when available and falls back to the all-cause
definition when chart review has not been completed, allowing us to use as many
samples as we have available while taking physicians' expert opinions into
account where possible (@fig-flowchart B).
We trained ML models to classify (in the case of the IDSA definition) or predict
(in the case of the three other definitions) severity for each definition and
determined how well OTU-based models perform for each of them.

@@

To calculate a clinically-relevant NNS for these models, we computed the
confusion matrix at the 95th percentile of risk for each prediction model
(@tbl-risk).
We excluded the IDSA severity models as the IDSA severity scores were calculated
on the day of diagnosis, thus they are classification rather than prediction
problems.
Furthermore, IDSA severity scores do not correlate well with disease-related
adverse events which are a more salient outcome to prevent.
Among the models predicting severe outcomes, those trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition.
Multiplying the NNS of the OTU-based models by the estimated NNT of 
`r fdx_nnt` for fidaxomicin yields NNB values of `r allcause_fdx_nnb` 
for all-cause severity, `r attrib_fdx_nnb` for attributable severity,
and `r pragmatic_fdx_nnb` for pragmatic severity.
Thus, in a hypothetical scenario where these assumptions about fidaxomicin hold
true, between `r min_fdx_nnb` and `r max_fdx_nnb` patients would need to be
predicted to experience a severe outcome and be treated with fidaxomicin in
order for one patient to benefit.
As the NNS values were computed at the 95th percentile of risk (where 5% of
patients screened are predicted to experience severity), these NNB values mean
that `r min_screen` to `r max_screen` total CDI patients would need to be
screened by an OTU-based prediction model in order for one patient to benefit.
For comparison, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data extracted two days after diagnosis and from a
smaller set of manually curated variables, achieving precision values of 0.417
(NNS = `r ehr_nns`) for the EHR model and 0.167 (NNS = `r curated_nns`) for the
curated model at the 95th percentile of risk
[@li_using_2019;@rao_clostridium_2015]. <!-- TODO possible to find NNS on day of diagnosis for EHR model-->
Pairing the prior EHR-based model with fidaxomicin would yield
an NNB of `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened for
one patient to benefit, although the EHR was extracted two days after diagnosis
while OTUs in this study are from stool samples collected on the day of diagnosis.
These estimates represent a proof-of-concept demonstration of the potential
value and trade-offs of deploying severity prediction models trained on
microbial factors versus EHRs to guide clinicians' treatment decisions.

@@

We trained ML models based on gut microbial communities on the day of CDI
diagnosis to predict CDI severity according to four different severity definitions.
The purpose of using the full datasets is to train the best models possible
given the constraints, while using the intersection dataset allows for comparing
severity definitions.
We found that models predicting pragmatic severity with as much data as
available performed best, and that performance dropped substantially when 
reducing to the intersection datasets for all definitions, likely due to the
particularly imbalanced nature of the all-cause and attributable definitions.
These results demonstrate the importance of using as many samples as possible when
data are sparse and the outcome is low prevalence, as well as the need to
incorporate physician's expertise when possible.

@@

However, it is not enough for models to perform well to justify deploying them
in a clinical setting; benefit over current practices must be shown [@wiens_no_2019].
Although no known treatment options are proven to reduce the risk of severe CDI
outcomes, fidaxomicin is promising due to its improved time to resolution and
reduced recurrence.
Fidaxomicin is also attractive as a preferential antibiotic option as
vancomycin-resistant enterococcus is on the rise and enterococci are known to
worsen CDI [@poduval_clostridium_2000;@smith_enterococci_2022].
We extended our analysis of clinical value to incorporate the number needed to
treat for fidaxomicin alongside the predictive models in order to calculate the
number needed to benefit.
The NNB contextualizes model performance within clinical reality, as it combines
both model performance and treatment performance [@liu_number_2019].
A more robust analysis of clinical value would further consider the cost of
treatment options versus the savings of averting severe outcomes, as economic
disparities are a major barrier to treatment in the US.
Cost-benefit analyses have found that fidaxomicin is more cost-effective than
vancomycin, largely due to the reduced risk of recurrence [@johnson_clinical_2021;@jiang_budget_2022].
While our analysis of clinical value is only a proof-of-concept, if evidence
emerges that new or existing treatments significantly reduce the risk of severe
CDI, our results can be incorporated into future considerations of whether to
build severity prediction models and what features should be incorporated.
In practice, EHR-based models are less costly to deploy than OTU-based models
and do not require additional clinical sample collection.
Amplicon sequencing is not typically performed for CDI patients, however,
routinely sequencing and profiling the microbial communities of CDI patients
could be justified if models that incorporate microbial features were shown to
improve patient outcomes.

In all, we found that our models to predict severity from features of the gut
microbiome performed moderately well.
Our approach enabled us to identify bacteria that contributed to model
performance and evaluate how well the state of the gut microbiome can predict
several different definitions of CDI severity.
Further work is needed to determine whether the performance of OTU-based models
is sufficient to justify their deployment in clinical settings, especially as
compared to EHR-based models.
If and when new evidence emerges of improved treatments to prevent severe CDI
outcomes, deploying performant and robust models for clinicians to tailor
treatment options promises to reduce morbidity and mortality of CDI.
<!--
we do not know which antibiotics were prescribed to treat these CDI cases, nor
which antibiotics patients may have taken prior to the CDI.
differences in microbiota between patients may be due to different abx...
different antibiotics have been shown to create different forms of dysbiotic
microbiota [@berkell_microbiota-based_2021] and
differential cdi clearance [@tomkovich_initial_2020].

>>>>

There is not currently a consensus definition of CDI severity.
Some scoring systems leverage clinical data available during the course of CDI,
while others focus on adverse outcomes of CDI at 30 days after diagnosis
[@ressler_defining_2021].
We explored four different ways to define CDI cases as severe or not (@fig-flowchart).
The "IDSA" definition of severe CDI is based on laboratory values collected on
the day of diagnosis, with a case being severe if serum creatinine level is
greater than or equal to $1.5 mg/dL$ and the white blood cell count is greater
than or equal to $15 k/\mu L$ [@mcdonald_clinical_2018].
However, the IDSA score is known to be a poor predictor of adverse outcomes
[@stevens_validation_2020], although it is straightforward to collect.
The remaining definitions we employed focus on the occurrence of adverse
outcomes, which may be more clinically relevant.
The "attributable" severity definition is based on disease-related complications
defined by the CDC, where an adverse event of ICU admission, colectomy, or death
occurs within 30 days of CDI diagnosis, and the adverse event is determined to
be attributable to the CDI by physician chart review [@mcdonald_recommendations_2007].
However, physician chart review is time-consuming and has not been completed for
all cases, so we defined "all-cause" severity where a case is severe if an
adverse event occurs within 30 days of the diagnosis regardless of the cause of
the adverse event.
Finally, we defined a "pragmatic" severity definition that makes use of the
attributable definition when available and falls back to the all-cause
definition when chart review has not been completed, allowing us to use as many
samples as we have available while taking physicians' expert opinions into
account where possible (@fig-flowchart B).
We trained ML models to classify (in the case of the IDSA definition) or predict
(in the case of the three other definitions) severity and determined how well
OTU-based models perform for each definition.

@@

To calculate a clinically-relevant NNS for these models, we computed the
confusion matrix at the 95th percentile of risk for each prediction model.
We excluded the IDSA severity models as the IDSA severity scores were calculated
on the day of diagnosis, thus they are classification rather than prediction
problems.
Furthermore, IDSA severity scores do not correlate well with disease-related
adverse events which are a more salient outcome to prevent.
Among the models predicting severe outcomes, those trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition (@tbl-risk).
Multiplying the NNS of the OTU-based models by the estimated NNT of 
`r fdx_nnt` for fidaxomicin yields NNB values of `r allcause_fdx_nnb` 
for all-cause severity, `r attrib_fdx_nnb` for attributable severity,
and `r pragmatic_fdx_nnb` for pragmatic severity.
Thus, in a hypothetical scenario where these assumptions about fidaxomicin hold
true, between `r min_fdx_nnb` and `r max_fdx_nnb` patients would need to be
predicted to experience a severe outcome and be treated with fidaxomicin in
order for one patient to benefit.
As the NNS values were computed at the 95th percentile of risk (where 5% of
patients screened are predicted to experience severity), these NNB values mean
that `r min_screen` to `r max_screen` total CDI patients would need to be
screened by an OTU-based prediction model in order for one patient to benefit.
For comparison, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data extracted two days after diagnosis and from a
smaller set of manually curated variables, achieving precision values of 0.417
(NNS = `r ehr_nns`) for the EHR model and 0.167 (NNS = `r curated_nns`) for the
curated model at the 95th percentile of risk
[@li_using_2019;@rao_clostridium_2015]. <!-- TODO possible to find NNS on day of diagnosis for EHR model-->
Pairing the prior EHR-based model with fidaxomicin would yield
an NNB of `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened for
one patient to benefit, although the EHR was extracted two days after diagnosis
while OTUs in this study are from stool samples collected on the day of diagnosis.
These estimates represent a proof-of-concept demonstration of the potential
value and trade-offs of deploying severity prediction models trained on
microbial factors versus EHRs to guide clinicians' treatment decisions.

@@

We trained ML models based on gut microbial communities on the day of CDI
diagnosis to predict CDI severity according to four different severity definitions.
The purpose of the full datasets was to train the best models possible given the
constraints, while using the intersection dataset allows for comparing severity
definitions.
We found that models predicting pragmatic severity with as much data as
available performed best, while models classifying IDSA severity outperformed
the all-cause and attributable definitions only with the inersection.
Performance dropped substantially when reducing to the intersection dataset for
all definitions, likely due to the particularly imbalanced nature of the
all-cause and attributable definitions.
These results demonstrate the importance of using as many samples as possible when
data are sparse and the outcome is low prevalence, as well as the need to
incorporate physician's expertise when possible.

@@

However, it is not enough for models to perform well to justify deploying them
in a clinical setting; benefit over current practices must be shown [@wiens_no_2019].
Although no known treatment options are proven to reduce the risk of severe CDI
outcomes, fidaxomicin is promising due to its improved time to resolution and
reduced recurrence.
Despite its increased cost, fidaxomicin is also attractive as a preferential
antibiotic option as vancomycin-resistant enterococcus is on the rise and
enterococci are known to worsen CDI
[@poduval_clostridium_2000;@smith_enterococci_2022].
We extended our analysis of clinical value to incorporate the number needed to
treat for fidaxomicin alongside the predictive models in order to calculate the
number needed to benefit.
The NNB contextualizes model performance within clinical reality, as it combines
both model performance and treatment effectiveness [@liu_number_2019].
A more robust analysis of clinical value would further consider the cost of
treatment options versus the savings of averting severe outcomes, as economic
disparities are a major barrier to treatment in the US [@johnson_clinical_2021].
Cost-benefit analyses based on clinical trial data have reported that
fidaxomicin may be as cost-effective as vancomycin as a treatment for initial
CDI cases, largely due to the reduced risk of recurrence
[@jiang_budget_2022;@reveles_fidaxomicin_2017].
While our analysis of clinical value is only a proof-of-concept, if evidence
emerges that new or existing treatments significantly reduce the risk of severe
CDI, our results can be incorporated into future considerations of whether to
build severity prediction models and what features should be incorporated.
In practice, EHR-based models are less costly to deploy than OTU-based models
and do not require additional clinical sample collection.
Amplicon sequencing is not typically performed for CDI patients, however,
routinely profiling the microbial communities of CDI patients could be justified
if models that incorporate microbial features were shown to improve patient
outcomes.

In all, we found that our models to predict severity from features of the gut
microbiome performed moderately well.
Our approach enabled us to identify bacteria that contributed to model
performance and evaluate how well the state of the gut microbiome can predict
several different definitions of CDI severity.
Further work is needed to determine whether the performance of OTU-based models
is sufficient to justify their deployment in clinical settings, especially as
compared to EHR-based models.
If and when new evidence emerges of improved treatments to prevent severe CDI
outcomes, deploying performant and robust models for clinicians to tailor
treatment options may improve patient outcomes and reduce the burden of severe
CDI.
<!--
we do not know which antibiotics were prescribed to treat these CDI cases, nor
which antibiotics patients may have taken prior to the CDI.
differences in microbiota between patients may be due to different abx...
different antibiotics have been shown to create different forms of dysbiotic
microbiota [@berkell_microbiota-based_2021] and
differential cdi clearance [@tomkovich_initial_2020].
